---
title: "Assignment04_823870359"
author: "Jake Kochmansky"
date: "2/23/2022"
output:
  html_document: default
---

```{r setup, include=TRUE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(ggplot2)
```

```{r Installing Data, echo = TRUE}
## Query URL:https://ed-data-portal.urban.org/data-explorer/explorer/?educationLevel=College&fipsNumbers=42&endYear=2019&timeframeType=most-recent&togglesList=23-530%2C22-528%2C66-452%2C66-451%2C66-450%2C6-1108%2C10-1120
EducationDataPortal_02.24.2022_years_since_entering_repay <- read.csv("~/Downloads/EducationDataPortal_02.24.2022_all_files/EducationDataPortal_02.24.2022_years_since_entering_repay.csv")
EducationDataPortal_02.24.2022_living_arrangement <- read.csv("~/Downloads/EducationDataPortal_02.24.2022_all_files/EducationDataPortal_02.24.2022_living_arrangement.csv")
EducationDataPortal_02.24.2022_datadictionary <- read.csv("~/Downloads/EducationDataPortal_02.24.2022_all_files/EducationDataPortal_02.24.2022_datadictionary.csv")
EducationDataPortal_02.24.2022_level_of_study.x.tuition_type <- read.csv("~/Downloads/EducationDataPortal_02.24.2022_all_files/EducationDataPortal_02.24.2022_level_of_study x tuition_type.csv")
EducationDataPortal_02.24.2022_award_level <- read.csv("~/Downloads/EducationDataPortal_02.24.2022_all_files/EducationDataPortal_02.24.2022_award_level.csv")
```
## Methodology
All data presented in the report was collected and sorted by the Urban Institute. For the data specially used below, the data can be found through the link in the accompanying code. The colleges highlighted below are limited to Pennsylvania for the most recent year data is available.

## Topic
With the student loan debt continuing to increase and pressure mounts on the Biden Administration to act, are their other options for students? The two most common counter arguments to the "cancelling" of student debt are: 1) these students chose to take on debt (i.e. there are other options); and 2) it is regressive (This topic will not be covered). Both appear to be fair criticisms of the policy. Let us look a little closer.

#1. Number of Colleges Offering Degree Types
If students are indeed choosing to take on debt, there would have to be other options. The most commonly cited alternative is attending community college that offers an Associate's degree. Is this a viable option? More institutions offer a Bachelor's degree than an Associate's degree, but with just under 100 institutions offering an Associate's degree in Pennsylvania, it appears that quantity of institutions is not an issue. NOTE: Penn State University has a plethora of satellite campuses that offer Associate's degrees which may skew this number up.
```{r # of Colleges that Offer Degree Type, echo=TRUE}
EducationDataPortal_02.24.2022_award_level %>%
  group_by(award_level) %>%
  count(awards) %>%
  ggplot() +
  geom_bar(
    mapping = aes(x = award_level, colour = award_level),
    show.legend = TRUE) +
    theme(axis.title.x=element_blank(),
      axis.text.x=element_blank(),
      axis.ticks.x=element_blank()) +
    labs(title = "Number of Colleges Offering Each Degree Type in PA", 
         y = "Number of Schools", 
         colour = "Degree/Award Type")
```
#2. Tuition Types
Most who have thought about attending a higher education have noticed the difference between in-state and out-of-state pricing. Based in the data presented below, we are unable to distinguish a large difference in the average of in-state and out-of-state tuition costs (NOTE: A hypothesis test has not been performed). However, practically, in-state vs. out-of-state can be thousands of dollars. Attending an institution in-state could save thousands, practically.

```{r Average Cost Per Tutition Type, echo=TRUE}
EducationDataPortal_02.24.2022_level_of_study.x.tuition_type %>%
  group_by(tuition_type,
           level_of_study = "Undergraduate") %>%
  summarise(mean_cost = mean(tuition_fees_ft, na.rm=TRUE),
            sd_cost = sd(tuition_fees_ft, na.rm = TRUE)) %>%
  ggplot()+
  geom_point(mapping = aes(x = tuition_type, y = mean_cost)) +
  labs(title = "Average Cost By Tuition Type",
                  x = "Tuition Type",
                  y = "Average Cost") +
  geom_errorbar(aes(x = tuition_type, y = mean_cost, ymin = mean_cost - sd_cost, ymax = mean_cost + sd_cost)
  ) 

```
#3. Book Costs
Necessary materials for class, such as books, are not included in typical pricing of higher education institutions. Unfortunately, students will not know the costs of books until after signing up for classes let alone before picking an institution, so there is does not lend to the argument of choice. Nonetheless, it appears that more than half of the institutions are below the mean. Transparency of book costs could lend to greater choice. 

```{r University Book Costs Across PA, echo = TRUE}
EducationDataPortal_02.24.2022_living_arrangement %>%
  group_by(inst_name) %>%
  summarise(mean_book_cost = mean(books_supplies)) %>%
  ggplot() +
  geom_density(mapping = aes(x = mean_book_cost, y = ..count..),
               fill = "gray") +
  geom_vline(mapping = aes(xintercept = mean(mean_book_cost)), 
             linetype = "dashed") +
  labs(title = "University Book Costs Across PA",
                  x = "Average Cost",
                  y = "Percentage of Institutions (/100")
```
#4. Repayment Rate by Family Income
One might be thinking how does repayment rate by family income contribute to choice? A student's ability to pay back a loan may seriously hinder their prospective choices. Family's with higher incomes may feel more comfortable taking a loan than low or mid-income families. Comparisons of the plots below show that higher income families tend to have larger repayment rates than lower income families at higher education institutions. This is no surprise and does not lend additional credence to the argument of choice.

```{r Comparing Reparyment Rate Per Family Income, echo= TRUE}

EducationDataPortal_02.24.2022_years_since_entering_repay %>%
  group_by(inst_name) %>%
  ggplot(aes(x = repay_rate_lowincome, y = factor(1)))+
  geom_point(alpha = .01, size = 5) +
  labs(y = NULL) +
  scale_x_discrete(breaks = 1:10 * 0.15) +
  scale_y_discrete(labels = NULL) +
  labs(title = "Repayment Rate For Low-Income Students",
       x = "Repayment Rate",
       y = NULL)
EducationDataPortal_02.24.2022_years_since_entering_repay %>%
  group_by(inst_name) %>%
  ggplot(aes(x = repay_rate_midincome, y = factor(1)))+
  geom_point(alpha = .01, size = 5) +
  labs(y = NULL) +
  scale_x_discrete(breaks = 1:10 * 0.15) +
  scale_y_discrete(labels = NULL) +
  labs(title = "Repayment Rate For Mid-Income Students",
       x = "Repayment Rate",
       y = NULL)
EducationDataPortal_02.24.2022_years_since_entering_repay %>%
  group_by(inst_name) %>%
  ggplot(aes(x = repay_rate_highincome, y = factor(1)))+
  geom_point(alpha = .01, size = 5) +
  labs(y = NULL) +
  scale_x_discrete(breaks = 1:10 * 0.15) +
  scale_y_discrete(labels = NULL) +
  labs(title = "Repayment Rate For High-Income Students",
       x = "Repayment Rate",
       y = NULL)
```
#Conclusion
The argument of choice in higher education as it relates to student debt is more nuanced than examined here. The supporting and contradictory data points presented do not fully capture the picture to make a conclusion on whether the argument of choice is sound. 

/end{document}
